Quasi-Monte Carlo Methods: What, Why, and How?

### 2025-01-24 HickernellMCQMC2024_QMCTutorial.tex
Updates from Overleaf

This tutorial describes  low discrepancy sequences  and their quality measures.  We demonstrate the performance gains possible with quasi-Monte Carlo methods.  Moreover, we describe how to formulate problems to realize the greatest performance gains using quasi-Monte Carlo methods.  We also briefly describe the use of quasi-Monte Carlo methods for problems beyond computing the mean, $\mu$.

@@

\section{Introduction} \label{sec:intro}
There are may settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity of rock, which affects the extraction of oil or gas, or
	\item Elementary particle interactions in a high energy physics experiment, which affect what is observed by detectors.
\end{itemize}
In such situations, the unknown inputs are often modeled using random vector variables or stochastic processes.  Computations are performed by generating a multitude of possible outcomes informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo (MC) method.

@@

\begin{equation}\label{eq:keisterA}
	\mu := \int_{\mathbb{R}^d} \cos(\lVert \bst \rVert_2) \exp(-\lVert \bst \rVert_2^2) \, \rmd \bst,
\end{equation}
where $\lVert \bst \rVert_2 := \sqrt{t_1^2 + \cdots + t_d^2}$.  This integral may be evaluated numerically by re-writing it in spherical coordinates as
\begin{equation}\label{eq:keisterExact}
	\mu = \frac{2 \pi^{d/2}}{\Gamma(d/2)}\int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r,
\end{equation}
where $2 \pi^{d/2}/\Gamma(d/2)$ is the surface area of the sphere in $d$ dimensions, and $\Gamma$ is the Gamma function.  The resulting one-dimensional integral is amenable to quadrature methods.  This non-trivial test case with a true value that can be easily calculated allows us to compute the numerical errors of various cubature schemes.

For the purpose of this illustration, we work with $\mu$ in its original form, \eqref{eq:keisterA} and approximate it by a sample mean,
\begin{equation} \label{eq:sample_mean}
	\hat{\mu}_n := \frac 1n \sum_{i=1}^n f(\bsx_i).
\end{equation}
This does require some preparation, which is discussed in generality in Section \ref{sec:reformulate}.

This $\mu$ is the expectation of $Y := g(\bsT) := \pi^{d/2} \cos(\lVert \bsT \rVert)$, where $T_1, \ldots, T_d$ are IID Gaussian (normal) random variables with zero mean and variance $1/2$, i.e., $\bsT :=(T_1, \ldots, T_d) \sim \calN(\bszero,\mathsf{I}/2)$:
\begin{equation}\label{eq:keisterB}
	\mu = \int_{\mathbb{R}^d} \underbrace{\pi^{d/2} \cos(\lVert \bst \rVert)}_{=:g(\bst)} \cdot \underbrace{\frac{\exp(-\lVert \bst \rVert^2)}{\pi^{d/2}}}_{\text{density of } \calN(\bszero,\mathsf{I}/2)} \, \rmd \bst = \bbE(Y) = \bbE[g(\bsT)].
\end{equation}
Virtually all LD sequences, which underlie qMC, are defined to approximate the standard uniform distribution, $\calU[0,1]^d$.  Thus, we perform a variable transformation $\bst = \bigl (\Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d) \bigr )/\sqrt{2}$,
where is $\Phi$ is the cumulative distribution function of the standard Gaussian random variable.  This reimagines the integral $\mu$ as the expectation of a function, $f$, of a standard uniform random variable:
\begin{multline}\label{eq:keisterC}
	\mu = \int_{[0,1]^d} \underbrace{\pi^{d/2} \cos\Bigl (\bigl \lVert \bigl( \Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d)\bigr) \bigr\rVert/\sqrt{2}  \Bigr)}_{=:f(\bsx)} \, \rmd \bsx \\
	= \bbE[Y]
	= \bbE[f(\bsX)], \qquad \bsX \sim \calU[0,1]^d.
\end{multline}

@@

\begin{figure}
	\centering
	\includegraphics[width =0.5\textwidth]{\figpath/n_is_16777216_d_is_6_n_rep_is_50_KeisterErrors.pdf}
	\caption{The relative error of approximating $\mu$ defined in \eqref{eq:keisterC} by the sample mean, $\hat{\mu}_n$, defined in \eqref{eq:sample_mean} for various choices of nodes, $\bsx_0, \bsx_1, \ldots$.  Grids have the largest error and LD nodes have the smallest error. The numerical computations used to generate this figure and the others in the is chapter may be reproduced using the Jupyter notebook \url{??}. \label{fig:keister-err}}
\end{figure}

@@

	Choosing the nodes to lie on a grid corresponds to a tensor product midpoint cubature rule, which would normally be expected to have an error of $\mathcal{O}(n^{-2/d})$  for general $d$.  The error decay of $\mathcal{O}(n^{-1/5})$ rather than $\mathcal{O}(n^{-2/6})$ for this example may be due to the lack of smoothness of $f$ near the boundaries of the unit cube.

	\item \emph{IID.}  Simple MC is a substantial improvement over grid nodes. As a reminder, for IID nodes the root mean squared error is
	\begin{equation}\label{eq:IIDerror}
		\sqrt{\mathbb{E}[(\mu - \hat{\mu})^2]} = \sqrt{\var(\hat{\mu})} = \sqrt{\frac{\var(f(\bsX))}{n}} = \frac{\mathrm{Std}(f(\bsX))}{n^{1/2}},
	\end{equation}
	where $\var$ denotes the variance and $\mathrm{Std}$ the standard deviation.  This $\mathcal{O}(n^{-1/2})$ decay is observed in Figure \ref{fig:keister-err}. For simple MC the sample size, $n$, can be any positive integer without affecting the rate of decay of the error.

@@

\begin{figure}
	\centering
	\includegraphics[width = 0.3\textwidth]{\figpath/64gridpts_d2.pdf}\qquad
	\includegraphics[width =  0.3\textwidth]{\figpath/64gridpts_d6.pdf}
	\caption{Although a two-dimensional grid covers the unit square rather well, a $d$-dimensional grid for modest $d$ does \emph{not} cover the unit cube well.  For example, one can only see four distinct points in a two dimensional projection of a six-dimensional grid with $64$ points. \label{fig:grid}}
\end{figure}

@@

One disadvantage of this construction is that it is not extensible. For the example in Figure \ref{fig:latticeconstruct} the first $8$ nodes do not fill the unit square well.  However, the set of nodes defined by even $i$ do a reasonable job.  This suggests a method for defining extensible lattice sequences that was proposed independently in \cite{Mai81a} and \cite{HicEtal00}.

The extensible lattice sequence is defined by the one-dimensional van der Corput sequence, $\{\phi_b(0), \phi_b(1), \ldots\}$. This involves the so-called radical inverse function and essentially reflects the digits of the integer $i$ in base $b$ about the decimal point. For example, $\phi_2(6) = \phi_2(110_{2}) = {}_20.011 = 3/8$.  In general,
\begin{multline} \label{eq:vdc}
	\phi_b(i_0 + i_1b + i_2 b^2 + \cdots ) := i_0 b^{-1} + i_1 b^{-2} + i_2 b^{-3} + \cdots \in [0,1)
	\\
	 \text{where } i_0, i_1, \ldots \in \{0,\ldots b-1\}.
\end{multline}
For all non-negative integers, $m$, the first $n = b^m$ nodes in the van der Corput sequence correspond to the evenly spaced nodes $\{0, b^{-m}, \ldots, 1 - b^{-m} \}$---albeit in a different order.

To construct an extensible lattice, we replace $i/n$ in \eqref{eq:latticepts} by $\phi_b(i)$ to get
\begin{equation} \label{eq:extensiblelattice}
	\bsx_i^{\text{lat}} =\phi_b( i)\bsh \pmod \bsone, \qquad i = 0, 1 , \ldots.
\end{equation}
This reordering of nodes from the original construction allows us to preserve the lattice structure for the first $b^m$ nodes for any non-negative integer $m$. That is, $\{\bsx_0, \ldots, \bsx_{b^m-1}\}$ is a closed under addition modulo $\bsone$.

Figure \ref{fig:extensiblelatticeconstruct} shows the first $4$, $8$, and $16$ nodes of an extensible lattice in base $2$ with the same generator as in Figure \ref{fig:latticeconstruct}.  The blue dots are a copy of the nodes in the plot to the left, and the the orange diamonds correspond to a shifted copy of the blue dots modulo $\bsone$.  For the left plot, the shift is $\phi_2(2)\bsh = (1,11)/4$ or equivalently $(0.25,0.75)$.  For the middle plot, the shift is $\phi_2(4)\bsh = (1,11)/8$ or equivalently $(0.125,0.375)$. For the right plot, the shift is $\phi_2(8)\bsh = (1,11)/16$  or equivalently $(0.0625,0.6875)$.

\begin{figure}
	\centering
	\includegraphics[width = \textwidth]{\figpath/16_extensible_lattice_construct_d2.pdf}
	\caption{An extensible lattice corresponding to  Figure \ref{fig:latticeconstruct}  with the nodes reordered using the van der Corput sequence in base $2$.  For each plot the blue nodes are a copy of the nodes to the left and the orange nodes are a shifted copy (modulo $\bsone$) of the blue nodes. \label{fig:extensiblelatticeconstruct}}
\end{figure}

For the example in Figure \ref{fig:extensiblelatticeconstruct}, increasing $n$ beyond $16$ repeats the original nodes because the generating vector, $\bsh$, contains integers all less than $16$.  Obtaining a truly extensible lattice sequence requires that $\bsh$ be a vector of generalized integers---essentially integers with infinite numbers of nonzero digits---as explained in \cite{HicNie03a}, where the existence of good generating vectors is also proved.  In practice, one searches computationally for a $\bsh$ that produces good set of lattice nodes of size $n = b^m$ for a range of $m$ \cite{HicEtal00}. The construction of great generating vectors for lattices has attracted a great deal of interest and is discussed further in Section \ref{sec:optLD}.

@@

Another family of LD sequences built upon the van der Corput sequence \eqref{eq:vdc} is digital sequences \cite{DicPil10a,Nie92}.  For simplicity, we restrict ourselves to $b = 2$ and let $\oplus$ denote binary digitwise addition modulo $2$ (also known as digitwise exclusive or), e.g., $3/8 \oplus 3/4 = {}_20.011 \oplus {}_20.110 = {}_20.101 = 5/8$.  A digital sequence is defined as
\begin{equation} \label{eq:digital}
	\bsx_i^{\text{dig}} := i_0 \bsx^{\text{dig}} _1 \oplus i_1 \bsx^{\text{dig}}_2 \oplus i_2 \bsx^{\text{dig}} _4 \oplus \cdots \in [0,1)^d \quad \text{for }
	i = i_0 + i_12 + i_2 2^2 + \cdots,
\end{equation}
where $\bsx^{\text{dig}}_1, \bsx^{\text{dig}}_2, \ldots$ are carefully chosen.  (This is the van  der Corput sequence in base $2$ for $d=1$ and $x_\ell = 2^{-\ell}$.) The node set $\{ \bsx_0, \ldots, \bsx_{2^m-1}\}$  for non-negative integer $m$ is called a digital net and is closed under $\oplus$.

If $x_{ij\ell}$ denotes the $\ell^{\text{th}}$ digit of the $j^{\text{th}}$ coordinate of $\bsx_i$, then the literature often defines generating matrices, $\mathsf{C}_j = (x_{ij\ell})_{\ell,i = 1}^{M,N}$ for $j = 1, \ldots, d$, where $M$ is the maximum number of bits in the expression for $\bsx_i$, say $52$, and $2^N$ is the maximum number of nodes that is intended to be generated.  The digital sequence can then be defined equivalently as
\begin{equation} \label{eq:digitalB}
\renewcommand{\arraystretch}{1.5}
	\begin{pmatrix} x^{\text{dig}}_{ij1} \\ x^{\text{dig}}_{ij2} \\ \vdots \end{pmatrix}
	= \mathsf{C}_j \begin{pmatrix} i_0 \\ i_1 \\ \vdots \end{pmatrix} \quad \pmod{2}, \qquad j = 1,\ldots, d, \ i = 0, 1, \ldots.
\end{equation}

@@

To illustrate the $t$-value, consider the following three dimensional digital net with $2^3$ (eight) nodes, i.e., $m = 3$:
\begin{multline} \label{eq:smallnet}
	\{(0, 0,   0),
	(0.5,   0.5,   0.5  ),
	(0.25,  0.75,  0.75 ),
	(0.75,  0.25,  0.25 ),
	(0.125, 0.625, 0.375),\\
	(0.625, 0.125, 0.875),
	(0.375, 0.375, 0.625),
	(0.875, 0.875, 0.125)\}.
\end{multline}
Figure \ref{fig:elementinterval} shows several two dimensional projections of these nodes and the two-dimensional boxes (rectangles) of the form \eqref{eq:element_box} with $\lVert\bsk\rVert_1 = 3$.  Most of the boxes contain their fair share of one node, but the box $[0,1/4) \times [0,1) \times [0,1/2)$ contains two nodes and the adjacent box, $[1/4,1/2) \times [0,1) \times [0,1/2)$ contains no nodes.  Thus, the $t$-value cannot be $0$.  However, the $t$-value is $1$ because all boxes of the form  \eqref{eq:element_box} with $\lVert\bsk\rVert_1 = 2$, i.e., a volume of $2^{-2} = 1/4$ contain a fair share of $2^{m- \lVert \bsk \rVert_1} = 2$ nodes.  The node set \eqref{eq:smallnet} is a $(1,3,3)$-net.

@@

Digital sequence generators can be found via number theory or numerical search \cite{KuoJoe08a}.  The earliest instance is due to Sobol' \cite{Sob67}.\NKNote{I added these two references which were previously left blank. Were you thinking of other papers to cite here?}

\subsection{Halton sequences} \label{sec:Halton}  While lattice sequences and digital sequences have preferred sample sizes, $n$, Halton sequences have no preferred sample size.  The Halton sequence is defined in terms of the van der Corput sequence
\begin{equation}\label{eq:Halton}
	\bsx^{\text{Hal}}_i = \bigl( \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \bigr), \qquad i = 0, 1, \ldots,
\end{equation}
where $b_1, \ldots, b_d$ is a choice of $d$ distinct prime numbers.  Often they are chosen as the first $d$ prime numbers.  Figure \ref{fig:Halton} shows two dimensional projections of Halton nodes in six dimensions. \NKNote{Suggestion: The Halton construction is an extensible sequence of LD nodes. However, should one know $n \in \mathbb{N}$ in advance, a finite LD node set called the Hammersley point set can be defined similarly as
\begin{equation}\label{eq:hammersley}
    \bsx_i^{\text{Ham}} = \left( \frac{i}{n}, \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \right) \quad i = 0, 1, \ldots, .
\end{equation}
}

@@

Most quasi-Monte Carlo constructions are studied based on their asymptotic behavior as \(n \to \infty\), with numerous results on asymptotic rates \cite{NIED1978QMCBook,NovWoz10a}. However, few methods address finding optimal node configurations for specific \(n\) and \(d\) which can be critical in practical applications where each function evaluation is costly. Of course, one possible strategy is to truncate asymptotically optimal sequences to the desired \(n\), however, as discussed above many of the standard LD constructions have a preferred number of nodes with truncations leading to imbalances in equidistribution. We can often do better optimizing directly for the target \(n\) and \(d\).

@@

When constructing LD nodes, it is essential to ensure not only a well-distributed set of nodes in the $d$-dimensional space but also in its lower-dimensional projections (recall the Cartesian grid from \ref{sec:intro}, Figure \ref{fig:grid}). \texttt{MatBuilder} \cite{paulin2022} is a software tool developed precisely for this task; to generate generating matrices for digital sequence LD nodes possessing excellent low dimensional uniformity. In this software, the solutions to integer linear programming problems are used to build the columns of generator matrices in a greedy manner for the LD digital sequence.

@@

\section{Randomization} \label{sec:random}
The LD constructions described in the previous section have so far been \emph{deterministic}. The sample mean, $\hat{\mu}_n$, does not change each time it is computed like it would for IID nodes. While determinism has some advantages, randomization has substantial advantages as well.
\begin{itemize}
	\item Randomization done correctly removes bias in the estimator.
	\item Replications of random estimators can facilitate error estimates for the sample mean.
	\item Often the application of interest requires transforming the LD nodes to mimic other distributions.  This is the case in the Keister example of Section \ref{sec:intro}, as seen in \eqref{eq:sample_mean_Keister} where the nodes, $\{\bsx_i\}_{i=0}^{n-1}$, are transformed to mimic a Gaussian distribution.

@@

[make a scramble of an image]

@@

Instead of explicitly searching for good generating vectors or matrices, one can instead sample a few random generating vectors or generating matrices and then prune out bad choices \cite{}.  Random generating vectors $\bsh$ take $h_j \sim \calU\{0,\dots,n-1\}$. Random generating matrices for base $2$ digital nets set $c_{ij\ell} \sim \calU\{0,1\}$ for $i<\ell$ while enforcing $c_{iji}=1$ and $c_{ij\ell}=0$ for $i>\ell$.

@@

Figure  \ref{fig:ctrKer} (right)  displays the discrepancy for a Sobol' LD digital net for several modest values of $d$ and for $n =  1, 2, 4, 8, 16, \ldots$.  The decay rate of nearly $\calO(n^{-1})$ is observed for small $d$ and can be anticipated for a bit larger $d$.  Thus, the decay of the discrepancy mirrors the convergence rate for LD sequences for the Keister example in Section \ref{sec:intro} and Figure \ref{fig:keister-err}.

>>>>

This tutorial describes  low discrepancy sequences  and their quality measures.  We demonstrate the performance gains possible with quasi-Monte Carlo methods.  Moreover, we describe how to formulate problems to realize the greatest performance gains using quasi-Monte Carlo.  We also briefly describe the use of quasi-Monte Carlo methods for problems beyond computing the mean, $\mu$.

@@

\section{Introduction} \label{sec:intro}
There are many settings where key underlying quantities that affect the outcome are unknown, e.g.,
\begin{itemize}
	\item Future market forces, which affect financial risk,
	\item The porosity field of rock, which affects the extraction of oil or gas, or
	\item Elementary particle interactions in a high energy physics experiment, which affect what is observed by detectors.
\end{itemize}
In such situations, the unknown inputs are often modeled using random vector variables or stochastic processes.  Computations are performed by generating a multitude of possible outcomes informed by the assumed probability distribution of the input. These are used to estimate the mean, quantiles, and/or probability distribution of the outcome.  This is the essence of the Monte Carlo (MC) method.

@@

\begin{equation}\label{eq:keisterA}
	\mu := \int_{\mathbb{R}^d} \cos(\lVert \bst \rVert_2) \exp(-\lVert \bst \rVert_2^2) \, \rmd \bst,
\end{equation}
where $\lVert \bst \rVert_2 := \sqrt{t_1^2 + \cdots + t_d^2}$.  This integral may be evaluated numerically by re-writing it in spherical coordinates as
\begin{equation}\label{eq:keisterExact}
	\mu = \frac{2 \pi^{d/2}}{\Gamma(d/2)}\int_{0}^{\infty} \cos(r) \exp(-r^2) \, r^{d-1} \rmd r,
\end{equation}
where $2 \pi^{d/2}/\Gamma(d/2)$ is the surface area of the sphere in $d$ dimensions, and $\Gamma$ is the Gamma function.  The resulting one-dimensional integral is amenable to quadrature methods.  This non-trivial test case with a true value that can be easily calculated allows us to compute the numerical errors of various cubature schemes and contrast their performance.

For the purpose of this illustration, we work with $\mu$ in its original form, \eqref{eq:keisterA} rather than \eqref{eq:keisterExact}, and approximate it by a sample mean,
\begin{equation} \label{eq:sample_mean}
	\hat{\mu}_n := \frac 1n \sum_{i=1}^n f(\bsx_i).
\end{equation}
This does require some preparation to identify a suitable $f$, which is discussed in generality in Section \ref{sec:reformulate}.

This $\mu$ in \eqref{eq:keisterA} is the expectation of $Y := g(\bsT) := \pi^{d/2} \cos(\lVert \bsT \rVert)$, where $T_1, \ldots, T_d$ are IID Gaussian (normal) random variables with zero mean and variance $1/2$, i.e., $\bsT :=(T_1, \ldots, T_d) \sim \calN(\bszero,\mathsf{I}/2)$:
\begin{equation}\label{eq:keisterB}
	\mu = \int_{\mathbb{R}^d} \underbrace{\pi^{d/2} \cos(\lVert \bst \rVert)}_{=:g(\bst)}  \underbrace{\frac{\exp(-\lVert \bst \rVert^2)}{\pi^{d/2}}}_{\text{density of } \calN(\bszero,\mathsf{I}/2)} \, \rmd \bst = \bbE(Y) = \bbE[g(\bsT)].
\end{equation}
Virtually all LD sequences, which underlie qMC, are defined to approximate the standard uniform distribution, $\calU[0,1]^d$.  Thus, we perform a variable transformation $\bst = \bigl (\Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d) \bigr )/\sqrt{2}$,
where is $\Phi$ is the cumulative distribution function of the standard Gaussian random variable.  This reimagines the integral $\mu$ as the expectation of a function, $f$, of a standard uniform random variable:
\begin{multline}\label{eq:keisterC}
	\mu = \int_{[0,1]^d} \underbrace{\pi^{d/2} \cos\Bigl (\bigl \lVert \bigl( \Phi^{-1}(x_1), \ldots, \Phi^{-1}(x_d)\bigr) \bigr\rVert/\sqrt{2}  \Bigr)}_{=:f(\bsx)} \, \rmd \bsx \\
	= \bbE[Y]
	= \bbE[f(\bsX)], \qquad \bsX \sim \calU[0,1]^d.
\end{multline}

@@

\begin{figure}
	\centering
	\includegraphics[width =0.5\textwidth]{\figpath/n_is_16777216_d_is_6_n_rep_is_50_KeisterErrors.pdf}
	\caption{The relative error of approximating $\mu$ defined in \eqref{eq:keisterC} by the sample mean, $\hat{\mu}_n$, defined in \eqref{eq:sample_mean} for various choices of nodes, $\bsx_0, \bsx_1, \ldots$.  Grids have the largest error and LD nodes have the smallest error. The numerical computations used to generate this figure and the others in this chapter may be reproduced using the Jupyter notebook \url{??}. \label{fig:keister-err}}
\end{figure}

@@

	Choosing the nodes to lie on a grid corresponds to a tensor product midpoint cubature rule, which would normally be expected to have an error of $\mathcal{O}(n^{-2/d})$  for general $d$.  The error decay of $\mathcal{O}(n^{-1/5})$ rather than $\mathcal{O}(n^{-2/6})$ for this example may be due to the lack of smoothness of $f$ near the boundaries of the unit cube.
\begin{figure}
	\centering
	\includegraphics[width = 0.3\textwidth]{\figpath/64gridpts_d2.pdf}\qquad
	\includegraphics[width =  0.3\textwidth]{\figpath/64gridpts_d6.pdf}
	\caption{Although a two-dimensional grid covers the unit square rather well, a $d$-dimensional grid for modest $d$ does \emph{not} cover the unit cube well.  For example, one can only see four distinct points in a two dimensional projection of a six-dimensional grid with $64$ points. \label{fig:grid}}
\end{figure}

	\item \emph{IID.}  Simple MC is a substantial improvement over grid nodes. For IID nodes the root mean squared error is
	\begin{equation}\label{eq:IIDerror}
		\sqrt{\mathbb{E}[(\mu - \hat{\mu})^2]} = \sqrt{\var(\hat{\mu})} = \sqrt{\frac{\var(f(\bsX))}{n}} = \frac{\mathrm{Std}(f(\bsX))}{n^{1/2}},
	\end{equation}
	where $\var$ denotes the variance and $\mathrm{Std}$ the standard deviation.  This $\mathcal{O}(n^{-1/2})$ decay is observed in Figure \ref{fig:keister-err}. For simple MC the sample size, $n$, can be any positive integer without affecting the rate of decay of the error.

@@

@@

One disadvantage of this construction is that it is not extensible. For the example in Figure \ref{fig:latticeconstruct}, the first $8$ nodes do not fill the unit square well.  However, the set of nodes defined by even $i$ do a reasonable job.  This suggests a method for defining extensible lattice sequences that was proposed independently in \cite{Mai81a} and \cite{HicEtal00}.

An extensible lattice sequence in one dimension is defined by the van der Corput sequence in base $b$, $\{\phi_b(0), \phi_b(1), \ldots\}$. This involves the so-called radical inverse function and essentially reflects the digits of the integer $i$ in base $b$ about the decimal point. For example, $\phi_2(6) = \phi_2(110_{2}) = {}_20.011 = 3/8$.  In general,
\begin{multline} \label{eq:vdc}
	\phi_b(i_0 + i_1b + i_2 b^2 + \cdots ) := i_0 b^{-1} + i_1 b^{-2} + i_2 b^{-3} + \cdots \in [0,1)
	\\
	 \text{where } i_0, i_1, \ldots \in \{0,\ldots b-1\}.
\end{multline}
For all non-negative integers, $m$, the first $n = b^m$ nodes in the van der Corput sequence correspond to the evenly spaced nodes $\{0, b^{-m}, \ldots, 1 - b^{-m} \}$---albeit in a different order.

To construct an extensible lattice, we replace $i/n$ in \eqref{eq:latticepts} by $\phi_b(i)$ to get
\begin{equation} \label{eq:extensiblelattice}
	\bsx_i^{\text{lat}} =\phi_b( i)\bsh \pmod \bsone, \qquad i = 0, 1 , \ldots.
\end{equation}
This reordering of nodes from the original construction allows us to preserve the lattice structure for the first $b^m$ nodes for any non-negative integer $m$. That is, $\{\bsx^{\text{lat}}_0, \ldots, \bsx^{\text{lat}}_{b^m-1}\}$ is a closed under addition modulo $\bsone$.

Figure \ref{fig:extensiblelatticeconstruct} shows the first $4$, $8$, and $16$ nodes of an extensible lattice in base $2$ with the same generator as in Figure \ref{fig:latticeconstruct}.  The blue dots are a copy of the nodes in the plot to the left, and the the orange diamonds correspond to a shifted copy of the blue dots modulo $\bsone$.  For the left plot, the shift is $\phi_2(2)\bsh = (1,11)/4=(0.25,0.75)$.  For the middle plot, the shift is $\phi_2(4)\bsh = (1,11)/8=(0.125,0.375)$. For the right plot, the shift is $\phi_2(8)\bsh = (1,11)/16=(0.0625,0.6875)$.

\begin{figure}
	\centering
	\includegraphics[width = \textwidth]{\figpath/16_extensible_lattice_construct_d2.pdf}
	\caption{An extensible lattice corresponding to  Figure \ref{fig:latticeconstruct}  with the nodes reordered using the van der Corput sequence in base $2$.  For each plot the blue dots are a copy of the nodes to the left and the orange diamonds are a shifted copy (modulo $\bsone$) of the blue dots. \label{fig:extensiblelatticeconstruct}}
\end{figure}

For the example in Figure \ref{fig:extensiblelatticeconstruct}, increasing $n$ beyond $16$ repeats the original nodes because the generating vector, $\bsh$, contains integers all less than $16$.  Obtaining a truly extensible lattice sequence requires that $\bsh$ be a vector of generalized integers---essentially integers with infinite numbers of nonzero digits---as explained in \cite{HicNie03a}, where the existence of good generating vectors for infinite lattice sequences is also proved.  In practice, one searches computationally for a $\bsh$ that produces good set of lattice nodes of size $n = b^m$ for a range of $m$ \cite{HicEtal00}. The construction of great generating vectors for lattices has attracted a great deal of interest and is discussed further in Section \ref{sec:optLD}.

@@

Another family of LD sequences built upon the van der Corput sequence \eqref{eq:vdc} is digital sequences \cite{DicPil10a,Nie92}.  For simplicity, we restrict ourselves to $b = 2$ and let $\oplus$ denote binary digitwise addition modulo $2$ (also known as digitwise exclusive or), e.g., $3/8 \oplus 3/4 = {}_20.011 \oplus {}_20.110 = {}_20.101 = 5/8$.  A digital sequence is defined as
\begin{equation} \label{eq:digital}
	\bsx_i^{\text{dig}} := i_0 \bsx^{\text{dig}} _1 \oplus i_1 \bsx^{\text{dig}}_2 \oplus i_2 \bsx^{\text{dig}} _4 \oplus \cdots \in [0,1)^d \quad \text{for }
	i = i_0 + i_12 + i_2 2^2 + \cdots,
\end{equation}
where $\bsx^{\text{dig}}_1, \bsx^{\text{dig}}_2, \ldots$ are carefully chosen.  (For $d=1$ and $x^{\text{dig}}_{2^m} = 2^{-m-1}$, this is the van  der Corput sequence in base $2$.) The node set $\{ \bsx^{\text{dig}}_0, \ldots, \bsx^{\text{dig}}_{2^m-1}\}$  for non-negative integer $m$ is called a digital net and is closed under $\oplus$.

If $x^{\text{dig}}_{ij\ell}$ denotes the $\ell^{\text{th}}$ digit of the $j^{\text{th}}$ coordinate of $\bsx^{\text{dig}}_i$, then the literature often defines generating matrices, $\mathsf{C}_j = \bigl(x^{\text{dig}}_{ij\ell}\bigr)_{\ell,i = 1}^{M,N}$ for $j = 1, \ldots, d$, where $M$ is the maximum number of bits in the expression for $\bsx^{\text{dig}}_i$, say $52$, and $2^N$ is the maximum number of nodes that is intended to be generated.  The digital sequence can then be defined equivalently as
\begin{equation} \label{eq:digitalB}
\renewcommand{\arraystretch}{1.5}
	\begin{pmatrix} x^{\text{dig}}_{ij1} \\ x^{\text{dig}}_{ij2} \\ \vdots \end{pmatrix}
	= \mathsf{C}_j \begin{pmatrix} i_0 \\ i_1 \\ \vdots \end{pmatrix} \quad \pmod{2}, \qquad j = 1,\ldots, d, \ i = 0, 1, \ldots.
\end{equation}

@@

To illustrate the $t$-value, consider the following $d=3$ dimensional digital net with $2^3$ (eight) nodes, i.e., $m = 3$:
\begin{multline} \label{eq:smallnet}
	\{(0, 0,   0),
	(0.5,   0.5,   0.5  ),
	(0.25,  0.75,  0.75 ),
	(0.75,  0.25,  0.25 ),
	(0.125, 0.625, 0.375),\\
	(0.625, 0.125, 0.875),
	(0.375, 0.375, 0.625),
	(0.875, 0.875, 0.125)\}.
\end{multline}
Figure \ref{fig:elementinterval} shows several two dimensional projections of these nodes and the two-dimensional boxes (rectangles) of the form \eqref{eq:element_box} with $\lVert\bsk\rVert_1 = 3$.  Most of the boxes contain their fair share of one node, but the box $[0,1/4) \times [0,1) \times [0,1/2)$ contains two nodes and the adjacent box, $[1/4,1/2) \times [0,1) \times [0,1/2)$ contains no nodes.  Thus, the $t$-value cannot be $0$.  However, the $t$-value is $1$ because all boxes of the form  \eqref{eq:element_box} with $\lVert\bsk\rVert_1 = 2$, i.e., a volume of $2^{-2} = 1/4$ contain a fair share of $2^{m- \lVert \bsk \rVert_1} = 2$ nodes.  The node set \eqref{eq:smallnet} is a $(1,3,3)$-net.

@@

Digital sequence generators can be found via number theory \cite[Chapter 8]{DicPil10a} or numerical search (see \cite{KuoJoe08a} and \cite[Chapter 10]{DicPil10a}).  The earliest instance is due to Sobol' \cite{Sob67}. %\NKNote{I added these two references which were previously left blank. Were you thinking of other papers to cite here?}

\subsection{Halton sequences} \label{sec:Halton}  While lattice sequences and digital sequences have preferred sample sizes, $n$, Halton sequences have no preferred sample size.  The Halton sequence is defined in terms of the van der Corput sequence
\begin{equation}\label{eq:Halton}
	\bsx^{\text{Hal}}_i = \bigl( \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \bigr), \qquad i = 0, 1, \ldots,
\end{equation}
where $b_1, \ldots, b_d$ is a choice of $d$ distinct prime numbers.  Often they are chosen as the first $d$ prime numbers.  Figure \ref{fig:Halton} shows two dimensional projections of Halton nodes in six dimensions. %\NKNote{Suggestion: 
The Halton construction is an extensible LD sequence. However, should one know $n \in \mathbb{N}$ in advance, a finite $d+1$-dimensional LD node set called the Hammersley point set can be defined similarly as
\begin{equation}\label{eq:hammersley}
    \bsx_i^{\text{Ham}} = \left( \frac{i}{n}, \phi_{b_1}(i), \ldots, \phi_{b_d}(i) \right) \quad i = 0, 1, \ldots, .
\end{equation}

@@

Most quasi-Monte Carlo constructions are studied based on their asymptotic behavior as \(n \to \infty\), with numerous results on asymptotic rates \cite{NIED1978QMCBook,NovWoz10a}. However, few methods address finding optimal node configurations for specific \(n\) and \(d\) which can be critical in practical applications where each function evaluation is quite costly. Of course, one possible strategy is to truncate asymptotically optimal sequences to the desired \(n\), however, as discussed above many of the standard LD constructions have a preferred number of nodes with truncations leading to imbalances in equidistribution. We can often do better optimizing directly for the target \(n\) and \(d\).

@@

When constructing LD nodes, it is essential to ensure not only a well-distributed set of nodes in the $d$-dimensional space but also in its lower-dimensional projections (recall the Cartesian grid from Section \ref{sec:intro}, Figure \ref{fig:grid}). \texttt{MatBuilder} \cite{paulin2022} is a software tool developed precisely for this task; to generate generating matrices for digital sequence LD nodes possessing excellent low dimensional uniformity. In this software, the solutions to integer linear programming problems are used to build the columns of generator matrices in a greedy manner for the LD digital sequence.

@@

\section{Randomization} \label{sec:random}
The LD constructions described in the previous section have so far been \emph{deterministic}. The sample mean, $\hat{\mu}_n$, does not change each time it is computed like it would for IID nodes. While determinism has advantages, randomization has substantial advantages as well.
\begin{itemize}
	\item Randomization done correctly removes bias in the estimator, $\hat{\mu}_n$.
	\item Replications of random estimators can facilitate error estimates for the sample mean.
	\item Often the application of interest requires transforming the LD nodes to mimic other distributions.  This is the case in the Keister example of Section \ref{sec:intro}, as seen in \eqref{eq:sample_mean_Keister} where the nodes, $\{\bsx_i\}_{i=0}^{n-1}$, are transformed to mimic a Gaussian distribution.

@@

@@

Instead of explicitly searching for good generating vectors or matrices, one can instead sample a few random generating vectors or generating matrices and then prune out bad choices \cite{}.  Random generating vectors for lattices take $\bsh \overset{\text{IID}}{\sim} \calU\{1,\dots,n-1\}^d$. Random generating matrices for base $2$ digital nets---as described in \eqref{eq:digitalB}---set $x^{\text{dig}}_{ij\ell} \overset{\text{IID}}{\sim} \calU\{0,1\}$ for $i<\ell$ while enforcing $x_{iji}=1$ and $x_{ij\ell}=0$ for $i>\ell$.

@@

Figure  \ref{fig:ctrKer} (right)  displays the discrepancy for a Sobol' LD digital net for several modest values of $d$ and for $n =  1, 2, 4, 8, 16, \ldots$.  The asymptotic decay rate of nearly $\calO(n^{-1})$ is observed for small $d$ and can be anticipated for a bit larger $d$.  Thus, the decay of the discrepancy mirrors the convergence rate for LD sequences for the Keister example in Section \ref{sec:intro} and Figure \ref{fig:keister-err}.
